The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps

### 2015-06-12 Manuscript/manuscript.tex
text edits, comments are embedded as (PCC: comment here). Didn't get to the discussion yet.

\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens}

\author[1]{Clinton B. Leach$^*$}
\author[1]{Colleen T. Webb}
\affil[1]{Department of Biology, Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO, USA}
\author[2]{Paul C. Cross}
\affil[2]{Northern Rocky Mountain Science Center, US Geological Survey, Bozeman, Montana, USA}
\affil[*]{Corresponding author: clint.leach@colostate.edu}

@@

\begin{abstract} 
The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.  However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.  In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the traits of the pathogen (virulence, transmission, environmental longevity) and the distribution of habitat quality in the host metapopulation.  This model suggests that for certain pathogens, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes in turn may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.

@@

We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}
where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).  
We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

Despite generally facilitating pathogen spread, for a range of pathogens, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (red shaded regions in Fig~\ref{highvlow}).  
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left of panels in Fig~\ref{highvlow}).
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

However, for pathogens falling between these extremes (e.g. with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_1.pdf}
\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.  Darker shading corresponds to more frequent endemic dynamics.  The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
\label{endemic}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_3.pdf}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).  At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.  However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
\label{popsizes}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.4]{figure/figure_5.pdf}
\centering
\caption{Boxplots showing the probability of infection (i.e. proportion of time occupied by infectious hosts) as a function of patch quality.  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show the effect of preferential selection of high quality habitat (a-c, $\xi_{im} = 0$,  no preference; d-f, $\xi_{im} = 0.5$, preference for high quality). The bottom row demonstrates the existence of an ecological trap on high quality habitat resulting from the fitness costs of increased disease risk. Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers. Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.}
\label{pinfection}
\end{figure}

\begin{figure}
\includegraphics[scale=1]{figure/figure_6.pdf}
\centering
\caption{Probability of host extinction as a function of pathogen environmental longevity and habitat preference, with black showing no habitat preference ($\xi_{im} = 0$) and red showing preference for high quality habitat ($\xi_{im} = 0.5$).  Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.  The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.}
\label{pext}
\end{figure}
\clearpage

>>>>

\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps}

\author[1]{Clinton B. Leach$^*$}
\author[1]{Colleen T. Webb}
\affil[1]{Department of Biology, Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO, USA}
\author[2]{Paul C. Cross}
\affil[2]{U.S. Geological Survey, Northern Rocky Mountain Science Center, Bozeman, Montana, USA}
\affil[*]{Corresponding author: clint.leach@colostate.edu}

@@

\begin{abstract} 
The distribution of habitat quality plays an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy (REF).  However, the costs and benefits of different quality habitat might be modulated, or even reversed, by the presence of an environmentally persistent pathogen.  The presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.  We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.  This model suggests that for certain (delete or explain what the certain refers to) pathogens, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.

PCC: abstract reads well. You could consider putting in a few more details. e.g. can you characterize the conditions/parameter space that creates the traps rather than just saying they may exist.

\end{abstract}

@@

Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others (PCC: may need latin names, could include chytrid? I think it is also potentially noteworthy that these pathogens also seem to be particularly notorious in wildlife....therefore this is a big deal).  This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.  This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}) (PCC: probably need a different reference for your first reference to "ecological trap").  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.

Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g. plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.  In this study, we explore how a pathogen's traits interact with the distribution of habitat quality to influence pathogen spread and host occupancy.  We specifically focus on the relative importance of high and low quality habitats to host persistence in the presence of pathogens with different levels of environmental longevity.

@@

We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.  Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
\begin{align}
C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
\label{connectivity}
\end{align}

PCC: not sure I understand why nothing is a function of time.

where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say d_ij is a distance here when 2 paragraphs down you set d_ij = 1).  Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat (PCC: while maintaining the same initial occupancy?).
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion (PCC: reword management actions that favor expansion) to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).  
We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).

To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).  We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.

@@

A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e. high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e. low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to persist (PCC: careful on word choice between persist and invade) and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).

For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).  (PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.  Essentially, I'm intersecting Fig. 1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2). This is a very different sort of take home message. This is just a change of spin compared to how you phrase the paragraph.)
In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}). 
Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e. at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).

However, for pathogens falling between these extremes (e.g. with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}). (PCC: Yes, may want to lead with this.)
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
Again, metapopulations with a lattice structure displayed similar patterns (Supplement).

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_1.pdf}
\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.  Darker shading corresponds to more frequent endemic dynamics.  The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics. PCC notes: need words and symbols on the axes. need to label the colorbar. I'd also draw some arrows across the rows and columns to show how pathogen longevity increases from left to right and habitat quality varies across rows. I'd also remove all the whitespace btwn plots so that the they are not super small in the paper. You could also directly label on the plots above and below the shading host vs. pathogen extinction.  I would not refer to host extinction as "pandemics" as these are not synonyms.}
\label{endemic}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.5]{figure/figure_3.pdf}
\centering
\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.  Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.  Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).  At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.  However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.  PCC: took me a long time to understand this non-traditional boxplot. Personnally, I'd prefer a wide line for the inner quartiles and thin line for the outers. Then your eye is drawn towards and dwells on the most likely region rather than towards the unlikely region (the outer quartiles.)}
\label{popsizes}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.4]{figure/figure_5.pdf}
\centering
\caption{Boxplots showing the probability of infection (i.e. proportion of time occupied by infectious hosts) as a function of patch quality.  Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show the effect of preferential selection of high quality habitat (a-c, $\xi_{im} = 0$,  no preference; d-f, $\xi_{im} = 0.5$, preference for high quality). The bottom row demonstrates the existence of an ecological trap on high quality habitat resulting from the fitness costs of increased disease risk. (PCC: delete sentence goes in the main text, but it also isn't clear to me how you derive this conclusion from the figure) Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers. Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$. PCC: you can scale y to be the same across all plots, which would allow you to drop your y axes on col 2 & 3, same goes for dropping the x axis on row 1 and squishing it together. Again, I would lable on the plot itself how longevity runs acroll the columns and rows indicate preference. If your main intent of the figure is to compare zero vs. high preference, then you should have those 2 cases on the same plot (maybe as two lines, dropping the boxplots) rather than making the reader make this connection across plots. The information in the boxplot seems to distract, rather than contribute, to your message here. }
\label{pinfection}
\end{figure}

\begin{figure}
\includegraphics[scale=1]{figure/figure_6.pdf}
\centering
\caption{Probability of host extinction as a function of pathogen environmental longevity and habitat preference, with black showing no habitat preference ($\xi_{im} = 0$) and red showing preference for high quality habitat ($\xi_{im} = 0.5$).  Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.  The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.  (PCC: seems like you could say this in text w/o using a figure since you have only 6 points here and only 2 of them are different....i.e. The probability of host extinction was ten times higher when there was a preference for high quality patches and longer environmental persistence (0.3 vs. 0.03))}
\label{pext}
\end{figure}
\clearpage
